---
title: "PICRUSt Analyses"
subtitle: "Bag-of-Genes"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```
For PICRUSt analysis: 

Use file KO_metagenome_out > pred_metagenome_unstrat_descrip.tsv to conduct the following tests/measures: 

Mean relative abundance in Vitamin D treatment 

Standard error of mean in Vitamin D treatment 

Mean relative abundance in placebo  

Standard error of mean in placebo 

GLM differential abundance analysis between treatment vs placebo (Patient=Fixed effect) 

- P-value 

- DR corrected q-value 


Use file EC_metagenome_out > pred_metagenome_unstrat_descrip to conduct the same analyses as described for A. 

Use file pathways_out > path_abun_unstrat_descrip to conduct the same analyses as in A.  
```

```{r data, include=FALSE, echo=FALSE, warning=F, error=F, message=F}
# load packages
source("code/PICRUSt/load_packages.R")

# get data
source("code/PICRUSt/get_cleaned_data.R")

theme_set(theme_bw())
# pal = "Sequential"
# scale_colour_discrete <-  function(palname=pal, ...){
#   scale_colour_brewer(palette=palname, ...)
# }
# scale_fill_discrete <-  function(palname=pal, ...){
#   scale_fill_brewer(palette=palname, ...)
# }

knitr::opts_chunk$set(out.width = "225%", out.height = 10)

options(digits=5, scipen=5)

# transform to relative abundances 
# in scale of 0-100
phylo.data.nci.umd <- transform_sample_counts(phylo.data.nci.umd, function(x){100*(x / sum(x))})

# melt data down for use
dat.16s <- phyloseq::sample_data(phylo.data.nci.umd)
```


# Part 1 EC Data

```{r pi-dat-ec}

pi.dat <- readr::read_tsv("analysis/PICRUSt/ec_pred_metagenome_unstrat_descrip.tsv")

# aggregate descriptions to get 1 description per row - unique;y defined.
pi.dat<- pi.dat %>%
  group_by(description) %>%
  summarise(across(`VDMT001-1`:`VDMT044-7`,.fns = sum))
# long format
pi.dat <- pi.dat %>%
  pivot_longer(
    cols=`VDMT001-1`:`VDMT044-7`,
    names_to = "ID",
    values_to = "Abundance"
  )

d <- as.data.frame(dat.16s)
mydata <- full_join(pi.dat, d)
mydata <- mydata %>%
  mutate(
    ID.n = as.numeric(as.factor(ID)),
    description.n = as.numeric(as.factor(description))
  ) %>%
  group_by(ID) %>%
  mutate(
    dem = ifelse(sum(Abundance)==0, 1, sum(Abundance)),
    RelAbundance = Abundance/dem*100) %>%
  ungroup()%>%
  group_by(description)%>%
  mutate(avgRA = mean(RelAbundance))

# only keep VDMT data
library("stringr")
mydata1 <- mydata[str_detect(mydata$ID, "VDMT0"),]

# added the cohort group (designates treatment vs placebo: 
randomization <- readr::read_excel("Randomization.xlsx", sheet = 1)

 # create participant ID column to match with randomization
mydata1$Participant <- substr(mydata1$ID, 1, 7)
mydata1 <- merge(mydata1, randomization, by='Participant', all.x=TRUE)

 # delete irrelevant columns 
empty_columns <- colSums(is.na(mydata1)) == nrow(mydata1)
mydata1 <- mydata1[, !empty_columns]
```


## Relative Abundance

```{r ec-ra}

d <- mydata1 %>%
  filter(description=="DNA-directed DNA polymerase")

t.test(d$RelAbundance ~ d$Cohort)

# Run on all descriptions
tb.ra1 <- mydata1 %>%
  group_by(description) %>%
  summarise(ng = n(),
            Overall.M = mean(RelAbundance),
            Overall.SE = sd(RelAbundance)/sqrt(ng))
tb.ra2m <- mydata1 %>%
  group_by(description, Cohort) %>%
  summarise(M = mean(RelAbundance)) %>%
  pivot_wider(id_cols = description,
              names_from = Cohort,
              values_from = M)
tb.ra2se <- mydata1 %>%
  group_by(description, Cohort) %>%
  summarise(ng=n(),SE = sd(RelAbundance)/sqrt(ng)) %>%
  pivot_wider(id_cols = description,
              names_from = Cohort,
              values_from = SE)
tb.ra2var <- mydata1 %>%
  group_by(description, Cohort) %>%
  summarise(ng=n(), VAR = var(RelAbundance)) %>%
  pivot_wider(id_cols = description,
              names_from = Cohort,
              values_from = VAR)
tb.ra2ng <- mydata1 %>%
  group_by(description, Cohort) %>%
  summarise(ng=n()) %>%
  pivot_wider(id_cols = description,
              names_from = Cohort,
              values_from = ng)
tb.ra <- left_join(tb.ra1, tb.ra2m)
tb.ra <- cbind(tb.ra, tb.ra2se[,-1])
tb.ra <- cbind(tb.ra, tb.ra2var[,-1])
tb.ra <- cbind(tb.ra, tb.ra2ng[,-1]) 

colnames(tb.ra) <- c("description", "ng", "Overall Mean", "Overall SE", "Placebo Mean", "Treatment Mean", "Placebo SE", "Treatment SE","Placebo Var", "Treatment Var", "Placebo Ng", "Treatment Ng")
tb.ra <- tb.ra %>%
  arrange(desc(`Overall Mean`))
tb.ra <- tb.ra[, c("description", "Overall Mean", "Overall SE","Treatment Mean","Treatment Var", "Treatment SE","Treatment Ng", "Placebo Mean","Placebo Var", "Placebo SE", "Placebo Ng")]

# compute t-test
tb.ra <- tb.ra %>%
  mutate(
    SEpooled = sqrt(`Treatment Var`/`Treatment Ng` + `Placebo Var`/`Placebo Ng`),
    t = (`Treatment Mean` - `Placebo Mean`)/(SEpooled),
    df = ((`Treatment Var`/`Treatment Ng` + `Placebo Var`/`Placebo Ng`)**2)/(((`Treatment Var`/`Treatment Ng`)**2)/(`Treatment Ng`-1) + ((`Placebo Var`/`Placebo Ng`)**2)/(`Placebo Ng`-1)),
    p = pt(q = abs(t), df=df, lower.tail = F)*2,
    fdr_p = p.adjust(p, method="fdr")
  )

kable(tb.ra, format="html", digits=5, caption="EC Data Average Relative Abundance of Each Description (sorted in descending order)") %>%
  kable_styling(full_width = T) %>%
  scroll_box(width = "100%", height="600px")

```


```{r ec-full-plot, warning=F, error=F, message=F, fig.dim=c(12,10), out.height="100%", out.width="100%"}


plot.dat <- tb.ra %>%
  arrange(desc(`Overall Mean`)) %>%
  slice_head(n=50)%>%
  mutate(
    description = fct_reorder(description, `Overall Mean`),
    description = factor(description, levels = levels(description), ordered=T)
  ) %>%
  arrange(description) %>%
  mutate(
    id = 1:n(),
    step = ifelse(id%%2 == 0, 1, 0),
    Mean_diff = `Treatment Mean` - `Placebo Mean`,
    diff_se = SEpooled,
    Mean_diff_ll = Mean_diff - qt(0.975, df)*diff_se,
    Mean_diff_ul = Mean_diff + qt(0.975, df)*diff_se
  ) %>%
  pivot_longer(
    cols=contains("Mean"),
    names_to = "group",
    values_to = "mean"
  )


p1.d <- plot.dat %>%
  filter(group %in% c("Treatment Mean","Placebo Mean")) %>%
  mutate(
    group = ifelse(group == "Treatment Mean", "Treatment", "Placebo"),
    col = ifelse(step == 1, "grey90", "white"),
    h=1, w=Inf
  )

p1 <- ggplot()+
  geom_tile(data = p1.d,
              aes(y = description, x=0,
                  height=h, width=w),
              fill = p1.d$col, color=p1.d$col)+
    geom_bar(data=p1.d,
             aes(x=mean, y=description,
                 group=group, color=group,
                 fill=group),
             stat="identity",position = "dodge",
             alpha = 1)+
    labs(x="Mean Proportion (%)")+
    theme_classic()+
    theme(
      legend.position = "bottom",
      plot.margin = unit(c(1,0,1,1), "lines")
    )

p2.d <- plot.dat %>%
  filter(group %in% c("Mean_diff", "Mean_diff_ll", "Mean_diff_ul")) %>%
  pivot_wider(
    names_from = group,
    values_from = mean
  ) %>%
  mutate(
    group = ifelse(Mean_diff > 0, "Treatment", "Placebo"),
    p = sprintf("%.3f", round(fdr_p,3)),
    ll = min(Mean_diff_ll)-0.01,
    ul = max(Mean_diff_ul)+0.01
  )
p2<-ggplot(p2.d, aes(x=Mean_diff, y=description))+
    geom_tile(data = p1.d,
              aes(y = description, x=0,
                  height=h, width=w),
              fill = p1.d$col, color=p1.d$col)+
    geom_vline(xintercept = 0, linetype="dashed", alpha=0.5)+
    geom_segment(aes(x=Mean_diff_ll, y=description, xend=Mean_diff_ul, yend=description))+
    geom_point(aes(fill=group, color=group))+
    geom_text(aes(label=p, x=unique(ul)+0.05))+
    coord_cartesian(xlim = c(unique(p2.d$ll), unique(p2.d$ul)),
                    clip = 'off') +
    annotate("text", x=unique(p2.d$ul)+0.08,y = 25,
             angle=90,
             label="q-value (FDR Corrected p-value)")+
    labs(x="Mean Difference in Proportions")+
    theme_classic()+
    theme(
      legend.position = "bottom",
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.line.y = element_blank(),
      axis.ticks.y = element_blank(),
      plot.margin = unit(c(1,4,1,0), "lines")
    )

# plot
p <- p1 + p2 +
  plot_annotation(title="EC Data: 50 most abundant descriptions")
p
ggsave("output/PICRUSt/picrust-EC.pdf",p,units="in", height=12, width=10)
```


## Modeling Difference between Treatment and Placebo 

For the modeling, we used a generalized linear mixed model (GLMM).

First, we looked at the biserial correlation between the abundance of each description and the tumor status.

```{r ec-corplot, warning=F, error=F, message=F, fig.dim=c(7,5), out.height="400px", out.width="100%"}

# to make cohort numeric so that the next model can use it 
cohort_nums <- mydata1$Cohort
cohort_nums[cohort_nums == "Treatment"] <- 1
cohort_nums[cohort_nums == "Placebo"] <- 2
mydata2 <- cbind(mydata1, cohort_nums)
mydata2$cohort_nums <- as.numeric(mydata2$cohort_nums)

tb <- mydata2 %>%
  group_by(description)%>%
  summarise(
    r = cor(cohort_nums, Abundance)
  ) %>%
  mutate(
    M=mean(r)
  )

ggplot(tb, aes(x=r))+
  geom_density()+
  geom_vline(aes(xintercept = M))+
  labs(x="Biserial Correlation",title="Relationship between description abundance and treatment (treatment vs. placebo)")+
  theme(panel.grid = element_blank())

```


Next, we need the data to be on an interpretable scale.
First, let's use the raw abundance like scale.

```{r ec-glm-overview, warning=F, error=F, message=F, fig.dim=c(7,5), out.height="400px", out.width="100%"}

p <- ggplot(mydata1, aes(x=Abundance))+
  geom_density()
p

mydata1 <- mydata1 %>%
  mutate(Abundance.dich=ifelse(Abundance==0, 0, 1))
table(mydata1$Abundance.dich)

```


Let's first run models by description so that we can avoid the nesting issue initially. We will come back to this to conduct the final model (it will be more powerful).

```{r ec-glm, message=F, eval=T}

DESCRIPTIONS <- unique(mydata1$description)
i<-1
dat0 <- mydata1 %>% filter(description==DESCRIPTIONS[i])

# quasipossion (approximately negative binom) give an approx. answer.
fit0 <- glm(
  Abundance ~ 1 + Cohort,
  data= dat0,
  family=quasipoisson(link = "log")
)
summary(fit0)

```

Next, let's model the percent relative abundance.
This will allow us to make inference about the difference average relative abundance.
Which is a simpler interpretation than trying to model differences in the log relative abundance which will need to be interpreted as the multiplicative relative change.

```{r ec-glm-approx}

results.out <- as.data.frame(matrix(ncol=4, nrow=length(DESCRIPTIONS)))
colnames(results.out) <- c("description", "Est", "SE", "p")
#results.out$description <-DESCRIPTIONS

# this for loop takes a very long time... 
i <- 1
for(i in 1:length(DESCRIPTIONS)){
#for(i in 1:5){ 
  dat0 <- mydata1 %>%
    filter(description == DESCRIPTIONS[i])%>%
    mutate(RelAbundance= RelAbundance*100)
  fit0 <- glm(
  RelAbundance ~ 1 + Cohort,
  data= dat0,
  family=quasipoisson(link = "log")
)
  fit.sum <- summary(fit0)
  results.out[i, 1] <- DESCRIPTIONS[i]
  results.out[i, 2:4] <- fit.sum$coefficients[2, c(1,2,4)]
}

results.out$fdr_p <- p.adjust(results.out$p, method="fdr")

kable(results.out, format="html", digits=3) %>%
  kable_styling(full_width = T)%>%
  scroll_box(width="100%", height="600px")


# merge with tb.ra for full results table
full.res <- left_join(tb.ra, results.out, by = "description")

write.csv(full.res, "output/PICRUSt/picrust_ec_data_results.csv", row.names = F)

```


# Part 2 KO Data

```{r pi-dat-ko}

pi.dat <- readr::read_tsv("analysis/PICRUSt/ko_pred_metagenome_unstrat_descrip.tsv")

# aggregate descriptions to get 1 description per row - unique;y defined.
pi.dat <- pi.dat %>%
  group_by(description) %>%
  summarise(across(`VDMT001-1`:`VDMT044-7`,.fns = sum))
# transform to long format
pi.dat <- pi.dat %>%
  pivot_longer(
    cols=`VDMT001-1`:`VDMT044-7`,
    names_to = "ID",
    values_to = "Abundance"
  )
d <- as.data.frame(dat.16s)
mydata <- full_join(pi.dat, d)

mydata <- mydata %>%
  mutate(
    ID.n = as.numeric(as.factor(ID)),
    description.n = as.numeric(as.factor(description))
  ) %>%
  group_by(ID) %>%
  mutate(
    dem = ifelse(sum(Abundance)==0, 1, sum(Abundance)),
    RelAbundance = Abundance/dem*100) %>%
  ungroup()%>%
  group_by(description)%>%
  mutate(avgRA = mean(RelAbundance))

# only keep VDMT data
mydata1 <- mydata[str_detect(mydata$ID, "VDMT0"),]

# create participant ID column to match with randomization
mydata1$Participant <- substr(mydata1$ID, 1, 7)
mydata1 <- merge(mydata1, randomization, by='Participant', all.x=TRUE)

#delete irrelevant columns 
empty_columns <- colSums(is.na(mydata1)) == nrow(mydata1)
mydata1 <- mydata1[, !empty_columns]

```


## Relative Abundance

```{r ko-ra}

tb.ra1 <- mydata1 %>%
  group_by(description) %>%
  summarise(ng = n(),
            Overall.M = mean(RelAbundance),
            Overall.SE = sd(RelAbundance)/sqrt(ng))
tb.ra2m <- mydata1 %>%
  group_by(description, Cohort) %>%
  summarise(M = mean(RelAbundance)) %>%
  pivot_wider(id_cols = description,
              names_from = Cohort,
              values_from = M)
tb.ra2se <- mydata1 %>%
  group_by(description, Cohort) %>%
  summarise(ng=n(),SE = sd(RelAbundance)/sqrt(ng)) %>%
  pivot_wider(id_cols = description,
              names_from = Cohort,
              values_from = SE)
tb.ra2var <- mydata1 %>%
  group_by(description, Cohort) %>%
  summarise(ng=n(), VAR = var(RelAbundance)) %>%
  pivot_wider(id_cols = description,
              names_from = Cohort, 
              values_from = VAR)
tb.ra2ng <- mydata1 %>%
  group_by(description, Cohort) %>%
  summarise(ng=n()) %>%
  pivot_wider(id_cols = description,
              names_from = Cohort,
              values_from = ng)
tb.ra <- left_join(tb.ra1, tb.ra2m)
tb.ra <- cbind(tb.ra, tb.ra2se[,-1])
tb.ra <- cbind(tb.ra, tb.ra2var[,-1])
tb.ra <- cbind(tb.ra, tb.ra2ng[,-1]) 

colnames(tb.ra) <- c("description", "ng", "Overall Mean", "Overall SE", "Placebo Mean", "Treatment Mean", "Placebo SE", "Treatment SE","Placebo Var", "Treatment Var", "Placebo Ng", "Treatment Ng")
tb.ra <- tb.ra %>%
  arrange(desc(`Overall Mean`))
tb.ra <- tb.ra[, c("description", "Overall Mean", "Overall SE","Treatment Mean","Treatment Var", "Treatment SE","Treatment Ng", "Placebo Mean","Placebo Var", "Placebo SE", "Placebo Ng")]

# compute t-test
tb.ra <- tb.ra %>%
  mutate(
    SEpooled = sqrt(`Treatment Var`/`Treatment Ng` + `Placebo Var`/`Placebo Ng`),
    t = (`Treatment Mean` - `Placebo Mean`)/(SEpooled),
    df = ((`Treatment Var`/`Treatment Ng` + `Placebo Var`/`Placebo Ng`)**2)/(((`Treatment Var`/`Treatment Ng`)**2)/(`Treatment Ng`-1) + ((`Placebo Var`/`Placebo Ng`)**2)/(`Placebo Ng`-1)),
    p = pt(q = abs(t), df=df, lower.tail = F)*2,
    fdr_p = p.adjust(p, method="fdr")
  )


kable(tb.ra, format="html", digits=5, caption="KO Data Average Relative Abundance of Each Description (sorted in descending order)") %>%
  kable_styling(full_width = T) %>%
  scroll_box(width = "100%", height="600px")

```

```{r ko-full-plot, warning=F, error=F, message=F, fig.dim=c(12,10), out.height="100%", out.width="100%"}


plot.dat <- tb.ra %>%
  arrange(desc(`Overall Mean`)) %>%
  slice_head(n=50)%>%
  mutate(
    description = fct_reorder(description, `Overall Mean`),
    description = factor(description, levels = levels(description), ordered=T)
  ) %>%
  arrange(description) %>%
  mutate(
    id = 1:n(),
    step = ifelse(id%%2 == 0, 1, 0),
    Mean_diff = `Treatment Mean` - `Placebo Mean`,
    diff_se = SEpooled,
    Mean_diff_ll = Mean_diff - qt(0.975, df)*diff_se,
    Mean_diff_ul = Mean_diff + qt(0.975, df)*diff_se
  ) %>%
  pivot_longer(
    cols=contains("Mean"),
    names_to = "group",
    values_to = "mean"
  )


p1.d <- plot.dat %>%
  filter(group %in% c("Treatment Mean","Placebo Mean")) %>%
  mutate(
    group = ifelse(group == "Treatment Mean", "Treatment", "Placebo"),
    col = ifelse(step == 1, "grey90", "white"),
    h=1, w=Inf
  )

p1 <- ggplot()+
  geom_tile(data = p1.d,
              aes(y = description, x=0,
                  height=h, width=w),
              fill = p1.d$col, color=p1.d$col)+
    geom_bar(data=p1.d,
             aes(x=mean, y=description,
                 group=group, color=group,
                 fill=group),
             stat="identity",position = "dodge",
             alpha = 1)+
    labs(x="Mean Proportion (%)")+
    theme_classic()+
    theme(
      legend.position = "bottom",
      plot.margin = unit(c(1,0,1,1), "lines")
    )
p2.d <- plot.dat %>%
  filter(group %in% c("Mean_diff", "Mean_diff_ll", "Mean_diff_ul")) %>%
  pivot_wider(
    names_from = group,
    values_from = mean
  ) %>%
  mutate(
    group = ifelse(Mean_diff > 0, "Treatment", "Placebo"),
    p = sprintf("%.3f", round(fdr_p,3)),
    ll = min(Mean_diff_ll)-0.01,
    ul = max(Mean_diff_ul)+0.01
  )
p2<-ggplot(p2.d, aes(x=Mean_diff, y=description))+
    geom_tile(data = p1.d,
              aes(y = description, x=0,
                  height=h, width=w),
              fill = p1.d$col, color=p1.d$col)+
    geom_vline(xintercept = 0, linetype="dashed", alpha=0.5)+
    geom_segment(aes(x=Mean_diff_ll, y=description, xend=Mean_diff_ul, yend=description))+
    geom_point(aes(fill=group, color=group))+
    geom_text(aes(label=p, x=unique(ul)+0.03))+
    coord_cartesian(xlim = c(unique(p2.d$ll), unique(p2.d$ul)),
                    clip = 'off') +
    annotate("text", x=unique(p2.d$ul)+0.05,y = 25,
             angle=90,
             label="q-value (FDR Corrected p-value)")+
    labs(x="Mean Difference in Proportions")+
    theme_classic()+
    theme(
      legend.position = "bottom",
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.line.y = element_blank(),
      axis.ticks.y = element_blank(),
      plot.margin = unit(c(1,4,1,0), "lines")
    )

# plot
p <- p1 + p2+
  plot_annotation(title="KO Data: 50 most abundant descriptions")
p
ggsave("output/PICRUSt/picrust-KO.pdf",p,units="in", height=12, width=10)
```

## Modeling Difference between Treatment and Placebo

For the modeling, we used a generalized linear mixed model (GLMM).

First, we looked at the biserial correlation between the abundance of each description and the tumor status.

```{r ko-corplot, warning=F, error=F, message=F, fig.dim=c(7,5), out.height="400px", out.width="100%"}

# to make cohort numeric
cohort_nums <- mydata1$Cohort
cohort_nums[cohort_nums == "Treatment"] <- 1
cohort_nums[cohort_nums == "Placebo"] <- 2
mydata2 <- cbind(mydata1, cohort_nums)
mydata2$cohort_nums <- as.numeric(mydata2$cohort_nums)

tb <- mydata2 %>%
  group_by(description)%>%
  summarise(
    r = cor(cohort_nums, Abundance)
  ) %>%
  mutate(
    M=mean(r)
  )

ggplot(tb, aes(x=r))+
  geom_density()+
  geom_vline(aes(xintercept = M))+
  labs(x="Biserial Correlation",title="Relationship between description abundance and treatment (treatment vs. placebo)")+
  theme(panel.grid = element_blank())

```


Next, we need the data to be on an interpretable scale.
First, let's use the raw abundance like scale.

```{r ko-glm-overview, warning=F, error=F, message=F, fig.dim=c(7,5), out.height="400px", out.width="100%"}

p <- ggplot(mydata1, aes(x=Abundance))+
  geom_density()
p

mydata1 <- mydata1 %>%
  mutate(Abundance.dich=ifelse(Abundance==0, 0, 1))
table(mydata1$Abundance.dich)

```


Let's first run models by description so that we can avoid the nesting issue initially. We will come back to this to conduct the final model (it will be more powerful).

```{r ko-glm, message=F, eval=T}

DESCRIPTIONS <- unique(mydata1$description)
i<-1
dat0 <- mydata1 %>% filter(description==DESCRIPTIONS[i])

# quasipossion (approximately negative binom) give an approx. answer.
fit0 <- glm(
  Abundance ~ 1 + Cohort,
  data= dat0,
  family=quasipoisson(link = "log")
)
summary(fit0)

```

Next, let's model the percent relative abundance.
This will allow us to make inference about the difference average relative abundance.
Which is a simpler interpretation than trying to model differences in the log relative abundance which will need to be interpreted as the multiplicative relative change.

```{r ko-glm-approx}

results.out <- as.data.frame(matrix(ncol=4, nrow=length(DESCRIPTIONS)))
colnames(results.out) <- c("description", "Est", "SE", "p")
#results.out$description <-DESCRIPTIONS

i <- 1
for(i in 1:length(DESCRIPTIONS)){
#for(i in 1:5){ 
  dat0 <- mydata1 %>%
    filter(description == DESCRIPTIONS[i])%>%
    mutate(RelAbundance= RelAbundance*100)
  fit0 <- glm(
  RelAbundance ~ 1 + Cohort,
  data= dat0,
  family=quasipoisson(link = "log")
)
  fit.sum <- summary(fit0)
  results.out[i, 1] <- DESCRIPTIONS[i]
  results.out[i, 2:4] <- fit.sum$coefficients[2, c(1,2,4)]
}

results.out$fdr_p <- p.adjust(results.out$p, method="fdr")

kable(results.out, format="html", digits=3) %>%
  kable_styling(full_width = T)%>%
  scroll_box(width="100%", height="600px")

# merge with tb.ra for full results table
full.res <- left_join(tb.ra, results.out, by = "description")

write.csv(full.res, "output/PICRUSt/picrust_ko_data_results.csv", row.names = F)

```


# Part 3 Pathways Data

```{r pi-dat-pathway}

pi.dat <- readr::read_tsv("analysis/PICRUSt/path_abun_unstrat_descrip.tsv")

# aggregate descriptions to get 1 description per row - uniquely defined.
pi.dat <- pi.dat %>%
  group_by(description) %>%
  summarise(across(`VDMT001-1`:`VDMT044-7`,.fns = sum))
# transform to long format
pi.dat <- pi.dat %>%
  pivot_longer(
    cols=`VDMT001-1`:`VDMT044-7`,
    names_to = "ID",
    values_to = "Abundance"
  )
d <- as.data.frame(dat.16s)
mydata <- full_join(pi.dat, d)

mydata <- mydata %>%
  mutate(
    ID.n = as.numeric(as.factor(ID)),
    description.n = as.numeric(as.factor(description))
  ) %>%
  group_by(ID) %>%
  mutate(
    dem = ifelse(sum(Abundance)==0, 1, sum(Abundance)),
    RelAbundance = Abundance/dem*100) %>%
  ungroup()%>%
  group_by(description)%>%
  mutate(avgRA = mean(RelAbundance))

# only keep VDMT data- I'm not even really sure why there is esoph data in mydata
mydata1 <- mydata[str_detect(mydata$ID, "VDMT0"),]

# create participant ID column to match with randomization
mydata1$Participant <- substr(mydata1$ID, 1, 7)
mydata1 <- merge(mydata1, randomization, by='Participant', all.x=TRUE)

#delete irrelevant columns 
empty_columns <- colSums(is.na(mydata1)) == nrow(mydata1)
mydata1 <- mydata1[, !empty_columns]
colnames(mydata1)
```


## Relative Abundance

```{r path-ra}

tb.ra1 <- mydata1 %>%
  group_by(description) %>%
  summarise(ng = n(),
            Overall.M = mean(RelAbundance),
            Overall.SE = sd(RelAbundance)/sqrt(ng))
tb.ra2m <- mydata1 %>%
  group_by(description, Cohort) %>%
  summarise(M = mean(RelAbundance)) %>%
  pivot_wider(id_cols = description,
              names_from = Cohort,
              values_from = M)
tb.ra2se <- mydata1 %>%
  group_by(description, Cohort) %>%
  summarise(ng=n(),SE = sd(RelAbundance)/sqrt(ng)) %>%
  pivot_wider(id_cols = description,
              names_from = Cohort,
              values_from = SE)
tb.ra2var <- mydata1 %>%
  group_by(description, Cohort) %>%
  summarise(ng=n(), VAR = var(RelAbundance)) %>%
  pivot_wider(id_cols = description,
              names_from = Cohort,
              values_from = VAR)
tb.ra2ng <- mydata1 %>%
  group_by(description, Cohort) %>%
  summarise(ng=n()) %>%
  pivot_wider(id_cols = description,
              names_from = Cohort,
              values_from = ng)
tb.ra <- left_join(tb.ra1, tb.ra2m)
tb.ra <- cbind(tb.ra, tb.ra2se[,-1])
tb.ra <- cbind(tb.ra, tb.ra2var[,-1])
tb.ra <- cbind(tb.ra, tb.ra2ng[,-1]) 

colnames(tb.ra) <- c("description", "ng", "Overall Mean", "Overall SE", "Placebo Mean", "Treatment Mean", "Placebo SE", "Treatment SE","Placebo Var", "Treatment Var", "Placebo Ng", "Treatment Ng")
tb.ra <- tb.ra %>%
  arrange(desc(`Overall Mean`))
tb.ra <- tb.ra[, c("description", "Overall Mean", "Overall SE","Treatment Mean","Treatment Var", "Treatment SE","Treatment Ng", "Placebo Mean","Placebo Var", "Placebo SE", "Placebo Ng")]

# compute t-test
tb.ra <- tb.ra %>%
  mutate(
    SEpooled = sqrt(`Treatment Var`/`Treatment Ng` + `Placebo Var`/`Placebo Ng`),
    t = (`Treatment Mean` - `Placebo Mean`)/(SEpooled),
    df = ((`Treatment Var`/`Treatment Ng` + `Placebo Var`/`Placebo Ng`)**2)/(((`Treatment Var`/`Treatment Ng`)**2)/(`Treatment Ng`-1) + ((`Placebo Var`/`Placebo Ng`)**2)/(`Placebo Ng`-1)),
    p = pt(q = abs(t), df=df, lower.tail = F)*2,
    fdr_p = p.adjust(p, method="fdr")
  )


kable(tb.ra, format="html", digits=5, caption="Pathway Data Average Relative Abundance of Each Description (sorted in descending order)") %>%
  kable_styling(full_width = T) %>%
  scroll_box(width = "100%", height="600px")

```


Let's try a nice bigger plot

```{r path-full-plot, warning=F, error=F, message=F, fig.dim=c(12,10), out.height="100%", out.width="100%"}


plot.dat <- tb.ra %>%
  arrange(desc(`Overall Mean`)) %>%
  slice_head(n=50)%>%
  mutate(
    description = fct_reorder(description, `Overall Mean`),
    description = factor(description, levels = levels(description), ordered=T)
  ) %>%
  arrange(description) %>%
  mutate(
    id = 1:n(),
    step = ifelse(id%%2 == 0, 1, 0),
    Mean_diff = `Treatment Mean` - `Placebo Mean`,
    diff_se = SEpooled,
    Mean_diff_ll = Mean_diff - qt(0.975, df)*diff_se,
    Mean_diff_ul = Mean_diff + qt(0.975, df)*diff_se
  ) %>%
  pivot_longer(
    cols=contains("Mean"),
    names_to = "group",
    values_to = "mean"
  )


p1.d <- plot.dat %>%
  filter(group %in% c("Treatment Mean","Placebo Mean")) %>%
  mutate(
    group = ifelse(group == "Treatment Mean", "Treatment", "Placebo"),
    col = ifelse(step == 1, "grey90", "white"),
    h=1, w=Inf
  )
p1 <- ggplot()+
  geom_tile(data = p1.d,
              aes(y = description, x=0,
                  height=h, width=w),
              fill = p1.d$col, color=p1.d$col)+
    geom_bar(data=p1.d,
             aes(x=mean, y=description,
                 group=group, color=group,
                 fill=group),
             stat="identity",position = "dodge",
             alpha = 1)+
    labs(x="Mean Proportion (%)")+
    theme_classic()+
    theme(
      legend.position = "bottom",
      plot.margin = unit(c(1,0,1,1), "lines")
    )
p2.d <- plot.dat %>%
  filter(group %in% c("Mean_diff", "Mean_diff_ll", "Mean_diff_ul")) %>%
  pivot_wider(
    names_from = group,
    values_from = mean
  ) %>%
  mutate(
    group = ifelse(Mean_diff > 0, "Treatment", "Placebo"),
    p = sprintf("%.3f", round(fdr_p,3)),
    ll = min(Mean_diff_ll)-0.01,
    ul = max(Mean_diff_ul)+0.01
  )
p2<-ggplot(p2.d, aes(x=Mean_diff, y=description))+
    geom_tile(data = p1.d,
              aes(y = description, x=0,
                  height=h, width=w),
              fill = p1.d$col, color=p1.d$col)+
    geom_vline(xintercept = 0, linetype="dashed", alpha=0.5)+
    geom_segment(aes(x=Mean_diff_ll, y=description, xend=Mean_diff_ul, yend=description))+
    geom_point(aes(fill=group, color=group))+
    geom_text(aes(label=p, x=unique(ul)+0.04))+
    coord_cartesian(xlim = c(unique(p2.d$ll), unique(p2.d$ul)),
                    clip = 'off') +
    annotate("text", x=unique(p2.d$ul)+0.08,y = 25,
             angle=90,
             label="q-value (FDR Corrected p-value)")+
    labs(x="Mean Difference in Proportions")+
    theme_classic()+
    theme(
      legend.position = "bottom",
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.line.y = element_blank(),
      axis.ticks.y = element_blank(),
      plot.margin = unit(c(1,4,1,0), "lines")
    )

# plot
p <- p1 + p2+
  plot_annotation(title="Path Data: 50 most abundant descriptions")
p
ggsave("output/PICRUSt/picrust-path.pdf",p,units="in", height=12, width=10)
```

## Modeling Difference between Treatment and Placebo

For the modeling, we used a generalized linear mixed model (GLMM).

First, we looked at the biserial correlation between the abundance of each description and the tumor status.

```{r path-corplot, warning=F, error=F, message=F, fig.dim=c(7,5), out.height="400px", out.width="100%"}

# Ella added to make cohort numeric
cohort_nums <- mydata1$Cohort
cohort_nums[cohort_nums == "Treatment"] <- 1
cohort_nums[cohort_nums == "Placebo"] <- 2
mydata2 <- cbind(mydata1, cohort_nums)
mydata2$cohort_nums <- as.numeric(mydata2$cohort_nums)

tb <- mydata2 %>%
  group_by(description)%>%
  summarise(
    r = cor(cohort_nums, Abundance)
  ) %>%
  mutate(
    M=mean(r)
  )

ggplot(tb, aes(x=r))+
  geom_density()+
  geom_vline(aes(xintercept = M))+
  labs(x="Biserial Correlation",title="Relationship between description abundance and treatment (treatment vs. placebo)")+
  theme(panel.grid = element_blank())

```



Next, we need the data to be on an interpretable scale.
First, let's use the raw abundance like scale.

```{r path-glmm-overview, warning=F, error=F, message=F, fig.dim=c(7,5), out.height="400px", out.width="100%"}

p <- ggplot(mydata1, aes(x=Abundance))+
  geom_density()
p

mydata1 <- mydata1 %>%
  mutate(Abundance.dich=ifelse(Abundance==0, 0, 1))
table(mydata1$Abundance.dich)


```


Let's first run models by description so that we can avoid the nesting issue initially. We will come back to this to conduct the final model (it will be more powerful).

```{r path-glm, message=F, eval=T}

DESCRIPTIONS <- unique(mydata1$description)
i<-1
dat0 <- mydata1 %>% filter(description==DESCRIPTIONS[i])

# quasipossion (approximately negative binom) give an approx. answer.
fit0 <- glm(
  Abundance ~ 1 + Cohort,
  data= dat0,
  family=quasipoisson(link = "log")
)
summary(fit0)

```

Next, let's model the percent relative abundance.
This will allow us to make inference about the difference average relative abundance.
Which is a simpler interpretation than trying to model differences in the log relative abundance which will need to be interpreted as the multiplicative relative change.

```{r path-glm-approx}


results.out <- as.data.frame(matrix(ncol=4, nrow=length(DESCRIPTIONS)))
colnames(results.out) <- c("description", "Est", "SE", "p")
#results.out$description <-DESCRIPTIONS

i <- 1
for(i in 1:length(DESCRIPTIONS)){
#for(i in 1:5){ 
  dat0 <- mydata1 %>%
    filter(description == DESCRIPTIONS[i])%>%
    mutate(RelAbundance= RelAbundance*100)
  fit0 <- glm(
  RelAbundance ~ 1 + Cohort,
  data= dat0,
  family=quasipoisson(link = "log")
)
  fit.sum <- summary(fit0)
  results.out[i, 1] <- DESCRIPTIONS[i]
  results.out[i, 2:4] <- fit.sum$coefficients[2, c(1,2,4)]
}

results.out$fdr_p <- p.adjust(results.out$p, method="fdr")

kable(results.out, format="html", digits=3) %>%
  kable_styling(full_width = T)%>%
  scroll_box(width="100%", height="600px")


# merge with tb.ra for full results table
full.res <- left_join(tb.ra, results.out, by = "description")

write.csv(full.res, "output/PICRUSt/picrust_path_data_results.csv", row.names = F)

```




# Misc. - disregarded for the VDMT analysis

```  
#{r stuff, eval=F}

library(brms)

# set up output object
results.out <- as.data.frame(matrix(ncol=7, nrow=length(DESCRIPTIONS)*3))
colnames(results.out) <- c("Description", "Parameter", "Est", "SE", "Q2.5", "Q97.5", "Rhat")
results.out$Description <- rep(DESCRIPTIONS, 3)

# for(i in 1:length(DESCRIPTIONS)){
for(i in 1:5){ 
  dat0 <- mydata1 %>%
    filter(description == DESCRIPTIONS[i])
  fit0 <- brms::brm(
    Abundance ~ 1 + Cohort,
    data= dat0,
    family=hurdle_lognormal(link="identity"),
    control = list(adapt_delta=0.99),
    silent = T)
  # the resulting "hu" parameter is the model for the proportion of zeros estimated
  #summary(fit0)
  #plot(fit0)
  # save the posterior summaries (intercept, tumor, and hu)
  fit.post <- round(posterior_summary(fit0),3)
  
  results.out[results.out$Description == DESCRIPTIONS[i], 2] <- c("b_Intercept", "b_treatment", "hu")
  results.out[results.out$Description == DESCRIPTIONS[i], 3:6] <- fit.post[c("b_Intercept", "b_treatment", "hu"),]
  results.out[results.out$Description == DESCRIPTIONS[i], 7] <- rhat(fit0)[c("b_Intercept", "b_treatment", "hu")]
  cat(".")
}

kable(results.out, format="html", digits=3) %>%
  kable_styling(full_width = T)%>%
  scroll_box(width="100%", height="600px")

```

Draft of a plots for these results

```
#{r glmm-plot1, eval=F}

res.out2 <- left_join(results.out, tb.ra,by=c("Description"="description")) %>% pivot_longer(
  cols = c("Tumor Mean", "Non-Tumor Mean"),
  names_to = "Tumor", values_to = "RelAbundance"
)

# based on above
dat0 <- res.out2 %>%
  mutate(estseR=abs(Est/SE))%>%
  na.omit()%>%
  filter(Parameter=="b_tumor", estseR >=0.25)

p <- ggplot(dat0, aes(y=Description, x=Est))+
  geom_point()+
  geom_segment(aes(y=Description, yend=Description, x=Q2.5, xend=Q97.5))
p

p2 <- ggplot(dat0, aes(x=RelAbundance, y=Description, color=Tumor))+
  geom_point(position = "dodge")
p2
```


Alternatively, we can model the relative abundance data.
For this, transform the RA data to be 0-100 instead of 0-1.

```
#{r glmm-ra, message=F, eval=F}


# set up output object
results.out <- as.data.frame(matrix(ncol=7, nrow=length(DESCRIPTIONS)*3))
colnames(results.out) <- c("Description", "Parameter", "Est", "SE", "Q2.5", "Q97.5", "Rhat")
results.out$Description <- rep(DESCRIPTIONS, 3)

# for(i in 1:length(DESCRIPTIONS)){
for(i in 1:5){ 
  dat0 <- mydata %>%
    filter(description == DESCRIPTIONS[i])%>%
    mutate(RelAbundance= RelAbundance*100)
  fit0 <- brms::brm(
    RelAbundance ~ 1 + tumor,
    data= dat0,
    family=hurdle_lognormal(link="identity"),
    control = list(adapt_delta=0.99),
    silent = T)
  # the resulting "hu" parameter is the model for the proportion of zeros estimated
  #summary(fit0)
  #plot(fit0)
  # save the posterior summaries (intercept, tumor, and hu)
  fit.post <- round(posterior_summary(fit0),3)
  
  results.out[results.out$Description == DESCRIPTIONS[i], 2] <- c("b_Intercept", "b_tumor", "hu")
  results.out[results.out$Description == DESCRIPTIONS[i], 3:6] <- fit.post[c("b_Intercept", "b_tumor", "hu"),]
  results.out[results.out$Description == DESCRIPTIONS[i], 7] <- rhat(fit0)[c("b_Intercept", "b_tumor", "hu")]
  cat(".")
}

kable(results.out, format="html", digits=3) %>%
  kable_styling(full_width = T)%>%
  scroll_box(width="100%", height="600px")

```

Draft of a plots for these results

```
#{r glmm-plot-ra, eval=F}

# based on above
p <- results.out %>%
  na.omit()%>%
  filter(Parameter=="b_tumor")%>%
  ggplot(aes(y=Description, x=Est))+
  geom_point()+
  geom_segment(aes(y=Description, yend=Description, x=Q2.5, xend=Q97.5))
p
```

For these analyses, we need to use some kind of hurdle/zero-inflated model to account for the large proportion of zeros in SOME of the the descriptions.
This creates a programming difficulty because only SOME really need this feature.
In the above, I tested two types of approaches.
First, we could model each descriptions separately (the quicker - time wise way).
Secondly, we could jointly model the descriptions using a multilevel model (2-level) hurdle model.
The latter is EXTREMELY computationally intensive.

```
#{r glmm-full, eval=F}

# random effects model (may be too large)
# let's try on a small subset

dat0 <- mydata %>%
  filter(description %in% DESCRIPTIONS[sample(1:length(DESCRIPTIONS),25)])
# priors
priors2 <-c(set_prior("lkj(2)", class = "cor"))
fit0 <- brms::brm(
  Abundance ~ 1 + tumor + (1 + tumor | description),
  prior = priors2,
  data= dat0,
  family=hurdle_lognormal(link="identity"), 
  #iter = 10000, warmup=2000,
  save_all_pars = T,
  control = list(adapt_delta=0.99))
# the resulting "hu" parameter is the model for the proportion of zeros estimated
summary(fit0)
plot(fit0)

```